Master Thesis

### 2013-06-26 acg.tex
Proofreading for publication

A \textbf{higher-order signature} is a set of elements that we call
\emph{constants}, each of which is associated with a type. Formally, it
is defined as a triple $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$,
where:
\begin{itemize}
  \item $C$ is the (finite) set of constants
  \item $A$ is a (finite) set of atomic types
  \item $\tau$ is the type associating mapping from $C$ to $\mathcal{T}(A)$,
    the set of types built over $A$
\end{itemize}

@@

\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they realized
  as.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$
  respectively). This property will be referred to as the
  \emph{homomorphism property} of a lexicon.
\end{itemize}

@@

\begin{itemize}
\item $t_3 = \synt{some}\ \synt{woman}$ \\ Mapping this term using the
  lexicon and $\beta$-reducing gives us $\mathcal{L}_{syntax}(t_3)
  =_{\beta} \lambda^{\circ} R.\ R\ (some\ +\ woman)$, which is a
  type-raised version of the string \emph{some woman}.
\item $t_4 = \lambda^{\circ}
  x.\ (\synt{some}\ \synt{woman})\ (\lambda^{\circ}
  y.\ \synt{love}\ x\ y)$ \\ Our verb phrase ends up being mapped onto
  a function which appends the phrase \emph{loves some woman} to its
  argument, $\mathcal{L}_{syntax}(t_4) =_{\beta} \lambda^{\circ}
  x.\ x\ +\ loves\ +\ some\ +\ woman$.
\item $t_5 = (\synt{every}\ \synt{man})\ (\lambda^{\circ}
  x.\ (\synt{some}\ \synt{woman})\ (\lambda^{\circ}
  y.\ \synt{love}\ x\ y))$ \\ Finally, the sentence (type $s$) is mapped
  into a simple $\textsc{String}$, $\mathcal{L}_{syntax}(t_5) =_{\beta}
  every\ +\ man\ +\ loves\ +\ some\ +\ woman$.
\end{itemize}

>>>>

A \textbf{higher-order signature} is a set of elements that we call
\emph{constants}, each of which is associated with a type. Formally, it
is defined as a triple $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$,
where:
\begin{itemize}
  \item $C$ is the (finite) set of constants
  \item $A$ is a (finite) set of atomic types
  \item $\tau$ is the type-associating mapping from $C$ to
    $\mathcal{T}(A)$, the set of types built over $A$
\end{itemize}

@@

\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they are
  realized as.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$
  respectively). This property will be referred to as the
  \emph{homomorphism property} of a lexicon.
\end{itemize}

@@

\begin{itemize}
\item $t_3 = \synt{some}\ \synt{woman}$ \\ Mapping this term using the
  lexicon and $\beta$-reducing gives us $\mathcal{L}_{syntax}(t_3)
  =_{\beta} \lambda^{\circ} R.\ R\ (some\ +\ woman)$, which is a
  type-raised version of the string \emph{some woman}.
\item $t_4 = \lambda^{\circ}
  x.\ (\synt{some}\ \synt{woman})\ (\lambda^{\circ}
  y.\ \synt{love}\ x\ y)$ \\ Our verb phrase ends up being mapped onto
  a function which appends the phrase \emph{loves some woman} to its
  argument, $\mathcal{L}_{syntax}(t_4) =_{\beta} \lambda^{\circ}
  x.\ x\ +\ loves\ +\ some\ +\ woman$.
\item $t_5 = (\synt{every}\ \synt{man})\ (\lambda^{\circ}
  x.\ (\synt{some}\ \synt{woman})\ (\lambda^{\circ}
  y.\ \synt{love}\ x\ y))$ \\ Finally, the sentence (type $S$) is mapped
  into a simple $\textsc{String}$, $\mathcal{L}_{syntax}(t_5) =_{\beta}
  every\ +\ man\ +\ loves\ +\ some\ +\ woman$.
\end{itemize}

### 2013-06-26 constraints.tex
Proofreading for publication

At a basic level, \emph{aucun} functions like any other determiner and
\emph{ne} as any verb modifier. However, we want to encode the
constraint that whenever a verb is modified by \emph{ne}, the modified
verb will demand that one of its arguments contains a negative
determiner. Since we need to make a distinction between phrases that
contain negative determiners and those that do not, we will need to
provide different types for both (two terms with the same type are
indistinguishable w.r.t. syntactic composition in a type-logical grammar
such as ACGs). Our type for \emph{aucun} will thus end up
being\footnote{Strictly following the EPTD for \emph{aucun} would lead
  us to give it a type $DET\_NEG{=}F$ and then give the two types
  $DET\_NEG{=}F \limp ((NP\_NEG{=}F \limp S) \limp S)$ and $DET\_NEG{=}T
  \limp ((NP\_NEG{=}T \limp S) \limp S)$ to nouns. In this
  demonstration, we prefer to align our examples with the treatment
  which is customary to ACG research.} $$N\_NEG{=}F \limp ((NP\_NEG{=}T
\limp S) \limp S)$$ instead of the usual $$N \limp ((NP \limp S) \limp
S)$$ The type for \emph{le} will need to be able handle cases where its
argument noun either already contains a negative determiner or not.

@@

However, this is not the only change we have to effect on our grammar in
order to properly handle the paired grammatical words for
negation. Consider the example sentence (\ref{ex:aucun-obj}) and the
parse tree assigned to it by Frigram on Figure~\ref{fig:parse-aucun}. It
is not enough that the embedded noun phrase \emph{aucun tatou} has a
type that tells us it contains a negative determiner. We need this bit
of internal information also at the level of the enclosing noun phrase
\emph{l'odeur d'aucun tatou} in which the former noun phrase is embedded
since this is a property of the enclosing noun phrase that is crucial to
its syntactic combinatorics. This presupposes the existence of a
mechanism for propagating this kind of information up the parse tree. In
the case of our ACG, this means that all the functions which modify
phrases must be aware whether or not their argument contains a negative
determiner and to convey this information in its result type as
well. Moreover, since the negative determiner can also come from a
complement of a prepositional phrase, the types of prepositions will
need to take into account the presence of negative determiners in both
their complements and the phrases they modify.

@@

There are two things that we would like to highlight about this
grammatical treatment. First off, we had to refine our types to make
them convey more than one piece of information in a single type
(e.g. something is a noun phrase and at the same time it is something
that contains a negative determiner). Second, a property that we wish to
express in the type of a term may be due to one of its subconstituents
and it is therefore necessary to encode the propagation of this property
up the parse tree in the types of all the intervening operators. In our
case, it meant that handling negative determiners and the paired
grammatical word for negation required us to not only define and type
constants for these two categories, but to also change the types of noun
modifiers (\emph{que ...}) and prepositions (\emph{de}).

@@

We also note that the splitting of the clause type $S$ into three finer
types $S\_EXT={NO}$, $S\_EXT{=}ROOT$ and $S\_EXT{=}EMB$ means that types
that work with $S$ regardless of its extraction status will have to
provide alternatives for all the possible variants.

>>>>

At a basic level, \emph{aucun} functions like any other determiner and
\emph{ne} as any other verb modifier. However, we want to encode the
constraint that whenever a verb is modified by \emph{ne}, the modified
verb will demand that one of its arguments contains a negative
determiner. Since we need to make a distinction between phrases that
contain negative determiners and those that do not, we will need to
provide different types for both (two terms with the same type are
indistinguishable w.r.t. syntactic composition in a type-logical grammar
such as ACGs). Our type for \emph{aucun} will thus end up
being\footnote{Strictly following the EPTD for \emph{aucun} would lead
  us to give it a type $DET\_NEG{=}T$ and then give the two types
  $DET\_NEG{=}F \limp ((NP\_NEG{=}F \limp S) \limp S)$ and $DET\_NEG{=}T
  \limp ((NP\_NEG{=}T \limp S) \limp S)$ to nouns. In this
  demonstration, we prefer to align our examples with the treatment
  which is customary in ACG research.} $$N\_NEG{=}F \limp ((NP\_NEG{=}T
\limp S) \limp S)$$ instead of the usual $$N \limp ((NP \limp S) \limp
S)$$ The type for \emph{le} will need to be able to handle cases where
its argument noun either already contains a negative determiner or not.

@@

However, this is not the only change we have to effect on our grammar in
order to properly handle the paired grammatical words for
negation. Consider the example sentence (\ref{ex:aucun-obj}) and the
parse tree assigned to it by Frigram on Figure~\ref{fig:parse-aucun}. It
is not enough that the embedded noun phrase \emph{aucun tatou} has a
type that tells us it contains a negative determiner. We need this bit
of internal information also at the level of the enclosing noun phrase
\emph{l'odeur d'aucun tatou}, in which the former noun phrase is
embedded, since this is a property of the enclosing noun phrase that is
crucial to its syntactic combinatorics. This presupposes the existence
of a mechanism for propagating this kind of information up the parse
tree. In the case of our ACG, this means that all the functions which
modify phrases must be aware whether or not their argument contains a
negative determiner and to convey this information in its result type as
well. Moreover, since the negative determiner can also come from a
complement of a prepositional phrase, the types of prepositions will
need to take into account the presence of negative determiners in both
their complements and the phrases they modify.

@@

There are two things that we would like to highlight about this
grammatical treatment. First off, we had to refine our types to make
them convey more than one piece of information in a single type
(e.g. something is a noun phrase and at the same time it is something
that contains a negative determiner). Second, a property that we wish to
express in the type of a term may be due to one of its subconstituents
and it is therefore necessary to encode the propagation of this property
up the parse tree in the types of all the intervening operators. In our
case, it meant that handling negative determiners and the paired
grammatical words for negation required us to not only define and type
constants for these two categories, but to also change the types of noun
modifiers (\emph{que ...}) and prepositions (\emph{de}).

@@

We also note that the splitting of the clause type $S$ into three finer
types $S\_EXT{=}NO$, $S\_EXT{=}ROOT$ and $S\_EXT{=}EMB$ means that types
that work with $S$ regardless of its extraction status will have to
provide alternatives for all the possible variants.

### 2013-06-26 gacg.tex
Proofreading for publication

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Because of the homomorphism property of lexicons, all
elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly, we have that $\mathcal{L}_{23}(A_2) \subseteq A_3$). These
new languages, which are formed by mapping an ACG abstract language
using a lexicon, are called \emph{object languages}. Their terms are
constrained not only by the types of the signature they are built upon
but also by the type system of the abstract signature.

@@

The definition of ACGs given above is quite practical in that it is
complete, i.e. it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{sec:acg}, where we defined ACGs as collections
of signatures associated with distinguished types and connected with
lexicons. Our definition only introduced the requisite machinery, but
delayed the delimitation of the way languages can be defined. Leaving
this question open will let us answer it now by providing an extension
of ACGs which will help us overcome the modularity issues that have been
the theme of Chapter~\ref{chap:constraints}.\footnote{Incidentally, this
  was not the reason for the way we presented ACGs in~\ref{sec:acg}. We
  chose this style as it grouped together the information in a more
  natural way. We approach signatures as descriptions of domains
  (strings, syntax, semantics), each one being associated with a
  distinguished type telling us which objects the domain has set out to
  describe (strings, sentences, truth values). Having distinguished
  types directly attached to signatures allows us to omit them from the
  ACG quadruples, $\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L},
  \cancel{S}\mathclose{>}$. However, this only leaves us with triples
  containing the two signatures and the lexicon connecting them, and
  since we can infer the signatures themselves from the domain and the
  range of the lexicon $\mathcal{L} : \Sigma_A \to \Sigma_O$, we do not
  have much motivation to introduce these tuples,
  $\mathopen{<}\cancel{\Sigma_A}, \cancel{\Sigma_O}, \mathcal{L},
  \cancel{S}\mathclose{>}$. Instead of composing ACGs, we compose just
  the lexicons.

On top of that, we will want to define signatures and lexicons to yield
multiple different object languages (at least one for the string
representations of forms and one for the semantic representations of
meaning), which would mean having at least two different ACGs which are
somehow linked together (e.g. by sharing the same abstract
signature). However, since these two grammars are so interrelated, we
would like to reason about them in a way that makes these relationships
explicit and thus easier to study as opposed to considering a collection
of grammars which are somehow implicitly similar.}

@@

First, we can show that our interpretation is sound meaning that any
arborescence adorned with signatures on the nodes, compatible lexicons
on the edges and a distinguished type on the root can only define ACG
languages (abstract or object languages). This is trivial, since the
language corresponding to the root is an ACG abstract language and the
class of ACG languages is closed under transformation by a lexicon and
therefore all of the languages corresponding to the descendants of the
root are ACG languages as well.

@@

In this section, we then examine the key properties and implications of
graphical ACGs.

@@

  \begin{itemize}
    \item For $v_1$, the case is trivial. $v_1$ has no predecessors and
      so $\mathcal{E}_{\mathcal{G}}(v_1) =
      \mathcal{I}_{\mathcal{G}}(v_1)$, which is an object language.
    \item For any other node $v_n$, we look at the definition of
      $\mathcal{E}_{\mathcal{G}}(v_n)$ and remark that the only
      operators are intersection and application of a lexicon with the
      operands being $\mathcal{I}_{\mathcal{G}}(v_n)$ (which is an
      object language) and the extrinsic languages of its predecessors
      (which can be constructed from object languages using
      intersections and lexicons thanks to the induction
      hypothesis). Therefore even $\mathcal{E}_{\mathcal{G}}(v_n)$ can
      be constructed from object languages using only intersections and
      lexicons.
  \end{itemize}

  Now we have to prove the converse. We do so by showing that extrinsic
  languages contain object languages and are closed under intersection
  and transformation by a lexicon.

@@

\begin{corollary}
  The set of extrinsic languages definable by graphical ACGs is the same
  as the set of object languages definable by ACGs if and only if the
  set of object languages definable by ACGs is closed under
  intersection.
\end{corollary}

@@

We turn to the problem of parsing with graphical ACGs. Parsing with ACGs
is a problem on the limits of decidability. When we consider ACGs whose
abstract signatures only contain types of order 2 or less, parsing is
possible in polynomial time for both linear \cite{salvati2005problemes}
and non-linear \cite{kanazawa2007parsing} type systems (such as the one
we introduced in~\ref{sec:acg}). If the order of the abstract signature
exceeds 2, the membership problem for linear ACGs becomes decidable if
and only if the satisfiability problem in the multiplicative exponential
fragment of linear logic is decidable \cite{de2004vector}, which is an
open problem. On the other hand, by extending the type system to include
dependent types, parsing can become undecidable too since in such a
system type checking itself is undecidable.

@@

  This definition allows us to directly establish the language of
  meanings which are expressible only by grammatical sentences in
  Figure~\ref{fig:both-patterns} and the language of morphological terms
  that have valid corresponding representations on all the other levels
  of linguistic description in Figure~\ref{fig:transducer-grammar}.
\item The diamond grammar of Figure~\ref{fig:diamond-grammar} is also no
  longer problematic since the labeling of nodes with terms guarantees
  that there is always a single term per node that can generate the term
  being parsed.
\end{itemize}

@@

  \item[$\mathcal{P}_{\mathcal{G}}(u) \subseteq
    \mathcal{E}_{\mathcal{G}}(u) \subseteq
    \mathcal{I}_{\mathcal{G}}(u)$] \hfill \\

>>>>

Further than that, we can look at the languages $\mathcal{L}_{12}(A_1)$
and $\mathcal{L}_{23}(A_2)$, images of the abstract languages given by
the lexicons. Because of the homomorphism property of lexicons, all
elements of $\mathcal{L}_{12}(A_1)$ will have the type
$\mathcal{L}_{12}(S_1) = S_2$ and will therefore also belong to $A_2$
(similarly, we have that $\mathcal{L}_{23}(A_2) \subseteq A_3$). These
new languages, which are formed by mapping an ACG abstract language
using a lexicon, are called \emph{object languages}. Their terms are
constrained not only by the types of the signature they are built upon
but also by the type system of the abstract signature and the lexicon
connecting the two.

@@

The definition of ACGs given above is quite practical in that it is
complete, i.e. it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{sec:acg}, where we defined ACGs as collections
of signatures associated with distinguished types and connected with
lexicons. Our definition only introduced the requisite machinery, but
delayed the delimitation of the ways languages can be defined. Leaving
this question open will let us answer it now by providing an extension
of ACGs which will help us overcome the modularity issues that have been
the theme of Chapter~\ref{chap:constraints}.\footnote{Incidentally, this
  was not the reason for the way we presented ACGs in~\ref{sec:acg}. We
  chose this style as it grouped together the information in a more
  natural way. We approach signatures as descriptions of domains
  (strings, syntax, semantics), each one being associated with a
  distinguished type telling us which objects the domain has set out to
  describe (strings, sentences, truth values). Having distinguished
  types directly attached to signatures allows us to omit them from the
  ACG quadruples, $\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L},
  \cancel{S}\mathclose{>}$. However, this only leaves us with triples
  containing the two signatures and the lexicon connecting them, and
  since we can infer the signatures themselves from the domain and the
  range of the lexicon $\mathcal{L} : \Sigma_A \to \Sigma_O$, we do not
  have much motivation to introduce these tuples,
  $\mathopen{<}\cancel{\Sigma_A}, \cancel{\Sigma_O}, \mathcal{L},
  \cancel{S}\mathclose{>}$. Instead of composing ACGs, we compose just
  the lexicons.

On top of that, we will want to define signatures and lexicons to yield
multiple different object languages (at least one for the string
representations of forms and one for the semantic representations of
meanings), which would mean having at least two different ACGs which are
somehow linked together (e.g. by sharing the same abstract
signature). However, since these two grammars are so interrelated, we
would like to reason about them in a way that makes these relationships
explicit and thus easier to study as opposed to considering a collection
of grammars which are somehow implicitly similar.}

@@

First, we can show that our interpretation is sound, meaning that any
arborescence adorned with signatures on the nodes, compatible lexicons
on the edges and a distinguished type on the root can only define ACG
languages (abstract or object languages). This is trivial, since the
language corresponding to the root is an ACG abstract language and the
class of ACG languages is closed under transformation by a lexicon and
therefore all of the languages corresponding to the descendants of the
root are ACG languages as well.

@@

In this section, we examine the key properties and implications of
graphical ACGs.

@@

  \begin{itemize}
    \item For $v_1$, the case is trivial. $v_1$ has no predecessors and
      so $\mathcal{E}_{\mathcal{G}}(v_1) =
      \mathcal{I}_{\mathcal{G}}(v_1)$, which is an object language.
    \item For any other node $v_n$, we look at the definition of
      $\mathcal{E}_{\mathcal{G}}(v_n)$ and remark that the only
      operators are intersection and application of a lexicon with the
      operands being $\mathcal{I}_{\mathcal{G}}(v_n)$ (which is an
      object language) and the extrinsic languages of its predecessors
      (which can be constructed from object languages using
      intersections and lexicons thanks to the induction
      hypothesis). Therefore, even $\mathcal{E}_{\mathcal{G}}(v_n)$ can
      be constructed from object languages using only intersections and
      lexicons.
  \end{itemize}

  Now we have to prove the converse. We do so by showing that extrinsic
  languages contain object languages and are closed on intersection and
  transformation by a lexicon.

@@

\begin{corollary}
  The set of extrinsic languages definable by graphical ACGs is the same
  as the set of object languages definable by ACGs if and only if the
  set of object languages definable by ACGs is closed on intersection.
\end{corollary}

@@

We now turn to the problem of parsing with graphical ACGs. Parsing with
ACGs is a problem on the limits of decidability. When we consider ACGs
whose abstract signatures only contain types of order 2 or less, parsing
is possible in polynomial time for both linear
\cite{salvati2005problemes} and non-linear \cite{kanazawa2007parsing}
type systems (such as the one we introduced in~\ref{sec:acg}). If the
order of the abstract signature exceeds 2, the membership problem for
linear ACGs becomes decidable if and only if the satisfiability problem
in the multiplicative exponential fragment of linear logic is decidable
\cite{de2004vector}, which is an open problem. On the other hand, by
extending the type system to include dependent types, parsing can become
undecidable too since in such a system type checking itself is
undecidable.

@@

  This definition allows us to directly establish the language of
  meanings which are expressible only by grammatical sentences in
  Figure~\ref{fig:both-patterns} and the language of morphological terms
  that have valid corresponding representations on all the other levels
  of linguistic description in Figure~\ref{fig:transducer-grammar}.
\item The diamond grammar of Figure~\ref{fig:diamond-grammar} is also no
  longer problematic since the labeling of nodes with terms guarantee
  that there is always a single term per node that can generate the term
  being parsed.
\end{itemize}

@@

  \item[The three languages are increasingly constraining:
    $\mathcal{P}_{\mathcal{G}}(u) \subseteq \mathcal{E}_{\mathcal{G}}(u)
    \subseteq \mathcal{I}_{\mathcal{G}}(u)$] \hfill \\
